---
title: "Workshop 8 TEMPLATE"
subtitle: "insert your section time here"
author: "insert your name here"
date: "insert your name here"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 6 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest")
```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# plant colors object 

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Eschscholzia californica" = esccal_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Encelia californica" = enccal_col,
  "Stipa pulchra" = stipul_col,
  "Grindelia camporum" = gricam_col
)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```

# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# making a new clean object from the original dataset 
drought_exp_clean <- drought_exp |> 
  # cleaning the column names for the new object 
  clean_names() |> 
  # updating the names in the species column from abbreviations to the full names
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
    )) |> 
  # Changing the water column to water treatment and spelling out the water treatment names
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"))  |> 
  # Reordering the species names within the data sat
  mutate(species_name = as_factor(species_name), # makes the data categorial
         species_name = fct_relevel(species_name, # reorders the data
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting the specific columns to only show in this new object
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: checking-structure

str(drought_exp_clean)
```

Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: ten-rows

# Displaying 10 random rows
slice_sample(drought_exp_clean, # dataframe to display 
             n = 10) # selecting 10 random rows 
```

# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 


```{r}
#| label: data-visualization-1-scatterplot

# Created ggplot base layer 
ggplot(data = drought_exp_clean,
       mapping = aes(x = sla, # x-axis is SLA 
                     y = total_g)) + # y-axis is total plant mass  
  # first layer is geom_point for all observations 
  geom_point(color = "dodgerblue3", # specifying color
             alpha = 0.7, # opacity
             shape = 21) + # shape
  labs(x = expression("Specific Leaf Area (cm"^2~"/g)"), 
       y = "Total Plant Mass (g)",
       title = "As SLA increases, plant mass (g) increases")
```


## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
#| label: data-visualization-2

# ggplot base layer 
ggplot(data = drought_exp_clean, 
       mapping = aes(x = water_treatment, # x-axis is water treatment
                     y = total_g, # y-axis is total plant mass
                     color = water_treatment)) + # color by treatment type 
  # first layer: geom_jitter
  geom_jitter(height = 0, #adjusting height and width of jitter points
              width = 0.1) + 
  # creating custom colors 
  scale_color_manual(values = c(
    "Well watered" = ww_col,
    "Drought stressed" = ds_col
  )) + 
  # adding custom labels 
  labs(x = "SLA",
       y = "Plant Mass (g)",
       title = "On average, well watered treatment had\nhigher plant mass than drought stressed treatment")
```

## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.

```{r}
#| label: data-visualization-3-jittplot

# base layer is ggplot
ggplot(data = drought_exp_clean,
       mapping = aes(x = species_name,
                     y = total_g,
                     color = species_name)) +
geom_jitter(height = 0,
            width = 0.1) + 
  #  Adding custom colors
  scale_color_manual(values = plant_colors) +
  # wrapping x-axis titles 
  scale_x_discrete(labels = label_wrap(10)) +
  # Adding custom titles 
  labs(x = "Species Name",
       y = "Total Plant Mass (g)",
       title = "Species differ in total plant mass") +
  # removing legend 
  theme(legend.position = "none")
```

# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model

model0 <- lm(total_g ~ 1, # formula = no predictors
             data = drought_exp_clean) # using clean data frame

# model 1: all predictors

model1 <- lm(
  total_g ~ sla + water_treatment + species_name, # formula all predictors
  data = drought_exp_clean) # using clean data frame 

# model 2: SLA and water treatment

model2 <- lm(
  total_g ~ sla + water_treatment, # formula only SLA and treatment
  data = drought_exp_clean # using clean data frame 
)

# model 3: SLA and species

model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean 
)

# model 4: water treatment and species

model4 <- lm(
  total_g ~ water_treatment + species_name, # formula: only treatment and species 
  data = drought_exp_clean # using clean drought data
)

# model 5: SLA

model5 <- lm(
  total_g ~ sla,
  data = drought_exp_clean)

# model 6: water treatment

model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)

# model 7: species

model7 <- lm(
  total_g ~ species_name,
  data = drought_exp_clean
)


```

## Model diagnostics

```{r}
#| label: check-diagnostics

#displaying plots in 2x2 grid
par(mfrow = c(2, 2))
# displaying model4 plots 
plot(model4)
```

**NOTES:***
Residuals appear normally distributed, homoscedastic (constant variance), and there are no clear outliers in the dataset that influence the model coefficient estimates 

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: calculating-aic

AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  arrange(AICc)
```

Insert code to display the summary for the best model. 

```{r}
#| label: model-summary 

summary(model4)
```

- intercept: model prediction for plant mass for Lotus scoparius in the drought stressed treatment.
- 


## Stop and think: what does this model mean?  

**What is the best model?**

[insert your response here]

**How much variation in the response (total mass, in grams) does this model explain?**

[insert your response here]

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

[insert your response here]

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.



# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.


# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







